Real Time Variational Density Forecasting

### 2017-03-13 Confirmation/confirmation_V2.tex
Python work

Electricity prices have been subject to an extensive forecasting literature and are known to be extremely volatile. For the 2011-2015 period in Victoria the mean price per megawatt hour (mWh) was \$42.06, with 99\% of observations less than \$100 per mWh, however on several occasions the price spiked to greater than $10,000\%$ of the mean value. For this reason market participants are increasingly interested in a predictive density forecast, as the probability that the price will rise above a particular threshold in the next time period can be more important than the mean value. To calculate a suitably accurate predictive density the Bayesian methodology is to be adopted in this thesis, as it accounts for both parameter and model uncertainty in a relatively straightforward manner, which can be difficult in frequentist methods. See \citet{Geweke2006} for a review on Bayesian forecasting and \citet{Gneiting2014} for density forecasting in general.

Bayesian methods are computationally complex as they require integration over the set of unknowns, whether parameters or latent stochastic variables. This set is typically quite large for models used to model short term electricity load and price as there are strong seasonal patterns in the data \citep{Taylor2003}. Analytical integration is typically not feasible to obtain the predictive distribution for such models, however there are a range of numerical and simulation-based techniques available for this purpose, notably Markov Chain Monte Carlo (MCMC) methods. While it is often possible to demonstrate theoretical convergence of MCMC algorithms, computation may be slow for even moderately complex models. This is problematic when attempting to forecast electricity load in Victoria, as realizations occur at five minute intervals, the predictive density for the next period's electricity load may not be available before it is observed.

@@

Mean Field Variational Bayes (see \citet{Bishop2006} for an overview) can be applied to exponetial family models with conjugate priors, and restricts the class of approximating distributions to the factorisable family. MFVB provides both the optimal form of the approximating distribution within this class, and a computationally simple coordinate ascent algorithm is available to optimise parameters.
Stochastic Variational Bayes (see \citet{Paisley2012} and \citet{Ranganath2014}) lifts both the restrictions for an exponential family model and the use of an approximating distribution that is factorisable. While parameter optimisation has a well developed gradient ascent algorithm, the determination of the functional form to be used is an active area of research.

We explore the use of a copula to augment the approximating distribution with a flexible dependence structure, which has seen success in \citet{Tran2015} and \citet{Ng2016}. The time series context of our implementation implies that MCMC samples for out of date samples are available, and we emphasise the use of this information to guide the construction of a vine copula so that the approximation has a dependence structure similiar to the true posterior.

@@

As discussed in Section \ref{sec:Intro}, a faster (albeit approximate) alternative to MCMC is VB. VB is an umbrella term for a broad collection of algorithms that introduce an approximating distribution $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ with the choice of the functional form $q$ and subsequent auxiliary parameter vector $\boldsymbol{\lambda}$ determined so that an error function is minimised. The classical choice for this error function is the Kullback-Leibler (KL) divergence \citep{Kullback1951} from $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ the true posterior $p(\boldsymbol{\theta} | y_{1:T})$. The KL divergence is defined by
\begin{equation}
\label{KL-def}
KL[q(\boldsymbol{\theta} | \boldsymbol{\lambda})\hspace{.1cm}||\hspace{.1cm}p(\boldsymbol{\theta} | y_{1:T})] = \int q(\boldsymbol{\theta} | \boldsymbol{\lambda}) \ln \left( \frac{q(\boldsymbol{\theta} | \boldsymbol{\lambda})}{p(\boldsymbol{\theta} |y_{1:T})}\right) d\boldsymbol{\theta}.
\end{equation}
The KL divergence is a non-negative, asymetric measure of the discrepancy between $p(\boldsymbol{\theta} | y_{1:T})$ and $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$  that will be equal to zero if and only if $p(\boldsymbol{\theta} | y_{1:T}) = q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ almost everywhere \citep{Bishop2006}. Note that $KL[q(\boldsymbol{\theta} | \boldsymbol{\lambda})\hspace{.1cm}||\hspace{.1cm}p(\boldsymbol{\theta} | y_{1:T})]$ can be expressed as
\begin{equation}
\label{KL-ELBO}
KL[q(\boldsymbol{\theta} | \boldsymbol{\lambda})\hspace{.1cm}||\hspace{.1cm}p(\boldsymbol{\theta} | y_{1:T})] = \ln(p(y_{1:T})) - \mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}), y_{1:T})
\end{equation}
where $\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}), y_{1:T})$ is referred to as the Evidence Lower Bound (ELBO), as it provides a lower bound on the unknown constant $\ln(p(y_{1:T}))$.  $\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}), y_{1:T})$ is defined by
\begin{equation}
\label{ELBO}
\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}), y_{1:T}) = \int q(\boldsymbol{\theta} | \boldsymbol{\lambda}) \ln (p(y_{1:T},\boldsymbol{\theta})) d\boldsymbol{\theta} -  \int q(\boldsymbol{\theta} | \boldsymbol{\lambda}) \ln (q(\boldsymbol{\theta} | \boldsymbol{\lambda})) d\boldsymbol{\theta}.
\end{equation}

Since $\ln(p(y_{1:T}))$ is constant with respect to $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$, maximising (\ref{ELBO}) with respect to $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ is equivalent to minimising (\ref{KL-def}). Maximising the ELBO is much more convenient than minimising the KL Divergence, as will be shown in the following sections in the context of the two major implementations of Variational Bayes we consider, Mean Field Variational Bayes and Stochastic Variational Bayes. Each of these implementations take advantage of the functional form of the ELBO, and offer computationally efficient algorithms to find a $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ that is optimal within a particular class of distributions. The statistical properties of $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ in MFVB and SVB are largely driven by the choice of the KL divergence from $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ to $p(\boldsymbol{\theta} | y_{1:T})$ as an error function, and there is significant interest in other error functions, such as the KL divergence from $p(\boldsymbol{\theta} | y_{1:T})$ to $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ \citep{Minka2001} and Stein's discrepency (\citealp{Liu2016}; \citealp{Ranganath2016b}).

@@

Mean Field Variational Bayes (MFVB) has origins in physics \citep{Peterson1987} and has since developed a large literature in machine learning, notably developed in \citet{Jordan1999} and \citet{Ghahramani2000}. MFVB restricts the class of distributions from which to select $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ to the set of factorisable distributions,
\begin{equation}
\label{mf1}
q(\boldsymbol{\theta} | \boldsymbol{\lambda}) = \prod_{i=1}^k q_i(\theta_i | \boldsymbol{\lambda_i}).
\end{equation}

Each of the $k$ components $\theta_i$ may be a vector, but in most implementations are scalars. Each $\theta_i$ has an associated vector $\boldsymbol{\lambda_i}$ which may have different dimension to $\theta_i$. $\boldsymbol{\lambda_i}$ is an auxillary parameter vector for the relevant factor $q_i$, which will be used in this section as shorthand notation for the distribution $q_i(\theta_i |\boldsymbol{\lambda_i})$.

@@

The expectations in (\ref{mf5}) - (\ref{mf8}) are available in a closed form for these distributions, and these are substituted in to derive the set of mean field equations:
\begin{align}
\tilde{\tau} &= \frac{\tilde{\beta}}{\tilde{\alpha}(T + \tau)} \label{mf9} \\
\tilde{\beta} &= \beta + 1/2\left((\tau + T)(\tilde{\gamma}^2 + \tilde{\tau}) - 2 \tilde{\gamma}\left(\sum_{t=1}^{T}y_t + \tau\right) + \sum_{t=1}^{T} y_t^2 + \tau \gamma^2 \right)\label{mf10}
\end{align}
$\tilde{\gamma}$ and $\tilde{\alpha}$ are omitted from the mean field equations as their values do not depend on other $\boldsymbol{\lambda}$ parameters.

@@

almost everywhere, where $\epsilon \sim U(0, 1)$ and $p^{-1}$ denotes the inverse CDF of the true posterior, will yeild $q(\boldsymbol{\theta} | \boldsymbol{\lambda} = p(\boldsymbol{\theta} | y_{1:T}$ and thus minimise the KL divergence. Attempts to model a suitably flexible function $f(\cdot, \cdot)$ include heirarchical models \citep{Ranganath2016a}, neural networks (\citealp{Kingma2014}; \citealp{Rezende2014}; \citealp{Rezende2015}; \citealp{Kingma2016}), and Gaussian processes \citep{Tran2016}.

These approaches introduce a large number of parameters that would considerably slow computation of updates, so we follow \citet{Tran2015} and use copulas to flexibly model approximations to the true posterior distribution. The copula modelling class provides a flexible way to construct these distributions, as they allow the dependence structure between parameters to be fit independently from the marginal distributions. \citet{Sklar1959} proves that any joint probability distribution can be written as the product of the marginals and a copula function,
\begin{equation}
\label{vc1}
p(\theta_1, \dots, \theta_k) = p(\theta_1) \dots p(\theta_k) c(P(\theta_1), \dots, P(\theta_k))
\end{equation}
where $p(\theta)$ is a pdf, $P(\theta)$ is a cdf, and $c(\cdot)$ is a copula.

@@

The likelihood of parameters $\boldsymbol{\theta} = (\phi_1, \phi_2, \sigma^2)'$, that is, the density of the observations $y_{1:T}$, given $\boldsymbol{\theta}$, viewed as a function of $\boldsymbol{\theta}$, is given by
\begin{align}
\label{likelihood}
L(\theta | y_{1:T}) &= p(y_1, y_2 | \theta) \prod_{t=3}^{T}p(y_t | y_{1:t-1}, \theta) \nonumber \\
&= \frac{1}{(2\pi)} |\Sigma|^{-1/2} \sigma^{-(T-2)} \exp \left\{ \frac{-1}{2} (y_1, y_2) \Sigma^{-1} (y_1, y_2)' \right\} \nonumber \\
&\times \exp \left\{ \frac{-1}{2\sigma^2} \left( \sum_{t=3}^{T}(y_t - \phi_1 y_{t-1} - \phi_2 y_{t-2})^2 \right) \right\}.
\end{align}

@@

The literature for electricity forecasting for both load and price is wide, and contains models based on many different fields, such as game theory, time-series modelling and neural networks, see \citet{Weron2014} for a recent review. Density forecasting for short term electricity load has had less attention, with \citet{Fan2012} and \citet{He2016} being notable examples, which used boot strapping and quantile regression respectively to generate forecast densities. Both of these approaches create forecast densities that are conditioned on the model and estimated parameters being correct, while the Bayesian approach used in this thesis can average over parameter and model uncertainty leading to a more accurate forecast density. This thesis also aims to re-estimate the model after each data point is observed to fully take advantage of any extra information on changes in parameters and thus further increase the accuracy of the forecast density.

>>>>

Electricity prices have been the subject of an extensive forecasting literature and are known to be extremely volatile. For the 2011-2015 period in Victoria the mean price per megawatt hour (mWh) was \$42.06, with 99\% of observations less than \$100 per mWh, however on several occasions the price spiked to greater than $10,000\%$ of the mean value. For this reason market participants are increasingly interested in a predictive density forecast, as the probability that the price will rise above a particular threshold in the next time period can be more important than the mean value. To calculate a suitably accurate predictive density the Bayesian methodology is to be adopted in this thesis, as it accounts for both parameter and model uncertainty in a relatively straightforward manner, which can be difficult in frequentist methods. See \citet{Geweke2006} for a review on Bayesian forecasting and \citet{Gneiting2014} for a discussion of density forecasting in general.

Bayesian methods are computationally complex as they require integration over the set of unknowns, whether parameters or latent stochastic variables. This set is typically quite large for models used to model short term electricity load and price as there are strong seasonal patterns in the data \citep{Taylor2003}. Analytical integration is typically not feasible to obtain the predictive distribution for such models, however there are a range of numerical and simulation-based techniques available for this purpose, notably Markov Chain Monte Carlo (MCMC) methods. While it is often possible to demonstrate theoretical convergence of MCMC algorithms, computation may be slow for even moderately complex models. This is problematic when attempting to forecast electricity load in Victoria, as realizations occur at five minute intervals, making it difficult to compute the predictive density for the next period's electricity load within the time frame.

@@

Mean Field Variational Bayes (see \citet{Bishop2006} for an overview) can be applied to exponetial family models with conjugate priors, and restricts the class of approximating distributions to the factorisable family. MFVB provides both the optimal form of the approximating distribution within this class, and a computationally simple coordinate ascent algorithm to optimise parameters.
Stochastic Variational Bayes (see \citet{Paisley2012} and \citet{Ranganath2014}) lifts both the restrictions for an exponential family model and the use of an approximating distribution that is factorisable. While parameter optimisation has a well developed gradient ascent algorithm, the determination of the functional form to be used is an active area of research.

We explore the use of a copula to augment the approximating distribution with a flexible dependence structure, which has seen success in \citet{Tran2015} and \citet{Ng2016}. The time series context of our implementation implies that MCMC samples for out of date samples are available, and we exploit this information in the construction of a vine copula so that the approximation has a dependence structure similiar to the true posterior.

@@

As discussed in Section \ref{sec:Intro}, a faster (albeit approximate) alternative to MCMC is VB. VB is an umbrella term for a broad collection of algorithms that introduce an approximating distribution $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ with the choice of the functional form $q$ and subsequent auxiliary parameter vector $\boldsymbol{\lambda}$ determined so that an error function is minimised. The classical choice for this error function is the Kullback-Leibler (KL) divergence \citep{Kullback1951} from $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ the true posterior $p(\boldsymbol{\theta} | y_{1:T})$. The KL divergence is defined by
\begin{equation}
\label{KL-def}
KL[q(\boldsymbol{\theta} | \boldsymbol{\lambda})\hspace{.1cm}||\hspace{.1cm}p(\boldsymbol{\theta} | y_{1:T})] = \int q(\boldsymbol{\theta} | \boldsymbol{\lambda}) \ln \left( \frac{q(\boldsymbol{\theta} | \boldsymbol{\lambda})}{p(\boldsymbol{\theta} |y_{1:T})}\right) d\boldsymbol{\theta}.
\end{equation}
The KL divergence is a non-negative, asymmetric measure of the discrepancy between $p(\boldsymbol{\theta} | y_{1:T})$ and $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$  that will be equal to zero if and only if $p(\boldsymbol{\theta} | y_{1:T}) = q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ almost everywhere \citep{Bishop2006}. Note that $KL[q(\boldsymbol{\theta} | \boldsymbol{\lambda})\hspace{.1cm}||\hspace{.1cm}p(\boldsymbol{\theta} | y_{1:T})]$ can be expressed as
\begin{equation}
\label{KL-ELBO}
KL[q(\boldsymbol{\theta} | \boldsymbol{\lambda})\hspace{.1cm}||\hspace{.1cm}p(\boldsymbol{\theta} | y_{1:T})] = \ln(p(y_{1:T})) - \mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}), y_{1:T})
\end{equation}
where $\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}), y_{1:T})$ is referred to as the Evidence Lower Bound (ELBO), as it provides a lower bound on the unknown constant $\ln(p(y_{1:T}))$.  $\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}), y_{1:T})$ is defined by
\begin{equation}
\label{ELBO}
\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}), y_{1:T}) = \int q(\boldsymbol{\theta} | \boldsymbol{\lambda}) \ln (p(y_{1:T},\boldsymbol{\theta})) d\boldsymbol{\theta} -  \int q(\boldsymbol{\theta} | \boldsymbol{\lambda}) \ln (q(\boldsymbol{\theta} | \boldsymbol{\lambda})) d\boldsymbol{\theta}.
\end{equation}

Since $\ln(p(y_{1:T}))$ is constant with respect to $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$, maximising (\ref{ELBO}) with respect to $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ is equivalent to minimising (\ref{KL-def}). Maximising the ELBO is much more convenient than minimising the KL Divergence, as will be shown in the following sections in the context of the two major implementations of Variational Bayes we consider, Mean Field Variational Bayes and Stochastic Variational Bayes. Each of these implementations takes advantage of the functional form of the ELBO, and offers a computationally efficient algorithm to for finding $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ that is optimal within a particular class of distributions. The statistical properties of $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ in MFVB and SVB are largely driven by the choice of the KL divergence from $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ to $p(\boldsymbol{\theta} | y_{1:T})$ as an error function, and there is significant interest in other error functions, such as the KL divergence from $p(\boldsymbol{\theta} | y_{1:T})$ to $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ \citep{Minka2001} and Stein's discrepency (\citealp{Liu2016}; \citealp{Ranganath2016b}).

@@

Mean Field Variational Bayes (MFVB) has its origin in physics \citep{Peterson1987} and has since developed a large literature in machine learning, notably \citet{Jordan1999} and \citet{Ghahramani2000}. MFVB restricts the class of distributions from which to select $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ to the set of factorisable distributions,
\begin{equation}
\label{mf1}
q(\boldsymbol{\theta} | \boldsymbol{\lambda}) = \prod_{i=1}^k q_i(\theta_i | \boldsymbol{\lambda_i}).
\end{equation}

The $k$ components $\theta_i$ may be a vector, but in most implementations are scalars. Each $\theta_i$ has an associated vector $\boldsymbol{\lambda_i}$ that may have a different dimension from $\theta_i$. $\boldsymbol{\lambda_i}$ is an auxillary parameter vector for the relevant factor $q_i$, which will be used in this section as shorthand notation for the distribution $q_i(\theta_i |\boldsymbol{\lambda_i})$.

@@

The expectations in (\ref{mf5}) - (\ref{mf8}) are available in a closed form for these distributions, and these are substituted into (\ref{mf6}) and (\ref{mf8}) to yield the set of mean field equations:
\begin{align}
\tilde{\tau} &= \frac{\tilde{\beta}}{\tilde{\alpha}(T + \tau)} \label{mf9} \\
\tilde{\beta} &= \beta + 1/2\left((\tau + T)(\tilde{\gamma}^2 + \tilde{\tau}) - 2 \tilde{\gamma}\left(\sum_{t=1}^{T}y_t + \tau\right) + \sum_{t=1}^{T} y_t^2 + \tau \gamma^2 \right).\label{mf10}
\end{align}

@@

almost everywhere, where $\epsilon \sim U(0, 1)$ and $p^{-1}$ denotes the inverse CDF of the true posterior, will yield $q(\boldsymbol{\theta} | \boldsymbol{\lambda}) = p(\boldsymbol{\theta} | y_{1:T})$ and thus minimise the KL divergence. Attempts to model a suitably flexible function $f(\cdot, \cdot)$ include heirarchical models \citep{Ranganath2016a}, neural networks (\citealp{Kingma2014}; \citealp{Rezende2014}; \citealp{Rezende2015}; \citealp{Kingma2016}), and Gaussian processes \citep{Tran2016}.

These approaches introduce a large number of parameters that would considerably slow computation of updates, so we follow \citet{Tran2015} and use copulas to flexibly model approximations to the true posterior distribution. The copula modelling class provides a flexible way to construct these distributions, as it allows the dependence structure between parameters to be fit independently from the marginal distributions. \citet{Sklar1959} proves that any joint probability distribution can be written as the product of the marginals and a copula function,
\begin{equation}
\label{vc1}
p(\theta_1, \dots, \theta_k) = p(\theta_1) \dots p(\theta_k) c(P(\theta_1), \dots, P(\theta_k))
\end{equation}
where $p(\theta)$ is a pdf, $P(\theta)$ is a cdf, and $c(\cdot)$ is a copula.

@@

The likelihood of parameters $\boldsymbol{\theta} = (\phi_1, \phi_2, \sigma^2)'$, that is, the density of the observations $y_{1:T}$, given $\boldsymbol{\theta}$, viewed as a function of $\boldsymbol{\theta}$, is given by
\begin{align}
\label{likelihood}
L(\theta | y_{1:T}) &= p(y_1, y_2 | \theta) \prod_{t=3}^{T}p(y_t | y_{1:t-1}, \theta) \nonumber \\
&\propto |\Sigma|^{-1/2} \sigma^{-(T-2)} \exp \left\{ \frac{-1}{2} (y_1, y_2) \Sigma^{-1} (y_1, y_2)' \right\} \nonumber \\
&\times \exp \left\{ \frac{-1}{2\sigma^2} \left( \sum_{t=3}^{T}(y_t - \phi_1 y_{t-1} - \phi_2 y_{t-2})^2 \right) \right\}.
\end{align}

@@

The literature for electricity forecasting for both load and price is wide, and contains models based on many different fields, such as game theory, time-series modelling and neural networks, see \citet{Weron2014} for a recent review. Density forecasting for short term electricity load has had less attention, with \citet{Fan2012} and \citet{He2016} being notable examples, which used bootstrapping and quantile regression respectively to generate forecast densities. Both of these approaches create forecast densities that are conditioned on the model and estimated parameters being correct, while the Bayesian approach used in this thesis can average over parameter and model uncertainty leading to a more accurate forecast density. This thesis also aims to re-estimate the model after each data point is observed to fully take advantage of any extra information on changes in parameters and thus further increase the accuracy of the forecast density.
